Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2019-08-13 README.Rmd
worked on procedure in paper
work from Tuesday before meeting

- Can I remove some of the sources from the references file that I am not going to use?

>>>>

- Paper:
    - Can I remove some of the sources from the references file that I am not going to use?
    - Can we talk about how best to present the LIME algorithm and some of the terms from the original paper that do not make sense in the context of the R package?
    - Subsections in the introduction?
    - See other comments in the paper
    - I am thinking about moving the paper to a new repository and then renaming this repository as LIME-research-journals and using it more as a place to try stuff out
- Graphics Group:
    - schedule
    - current list of volunteers
- Sandia Internship:
    - email from Danny

### 2019-08-13 journals/04-lime_algorithm_and_proposed_methods/04-lime_algorithm_and_proposed_methods.Rmd
work from Tuesday before meeting

---
title: "04-lime_algorithm_and_proposed_methods"
author: "Katherine Goode"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

@@

This journal contains a draft of the LIME algorithm and ideas for different binning methods that may be better than the ones currently implemented in the LIME package.

```{r}
# Load libraries

# Source functions
source("../../code/helper_functions.R")
```

# LIME Procedure

### Original Paper

@@

| Notation | Original Paper Definition | My Interpretation (in terms of a data table) |
| -------- | ------------------------- | ------------------------- |
| $x\in\mathbb{R}^d$ | Original representation of an instance being explained | Vector of observed features for one observation e.g. $(x_1, x_2,...,x_d)$ where $\textbf{X}$ is an $n$ by $d$ matrix | 
| $f:\mathbb{R}^d→\mathbb{R}$ | The model being explained...In classification, $f(x)$ is the probability (or a binary indicator) that $x$ belongs to a certain class. | This is a random forest in our case. |
| $x'\in\{0,1\}^{d'}$ | A binary vector for interpretable representation (where interpretable explanations are defined as "use a representation that is understandable by humans") | Varies based on the simulation method used - note that $1\le d' \le d$ since it is possible that feature selection may be performed to choose a smaller set of variables than originally in $x$ (One example: vector of indicator variables that indicate whether the observed value falls in the same bin as the instance being explained e.g. $x'_i = I[x_i\in \mbox{same variable } i \mbox{ bin as the instance being explained}]$) | In our example, this is the random forest model. |
|$G$ | A class of potentially interpretable models, such as linear models, decision trees, or falling rule lists |  <span style="color:blue"> I'm not sure what is considered $G$ in the R package - the fitting of the final ridge regression model or the ridge regression model used for feature selection? </span>|
| $g$ | An explanation as a model...can be readily presented to the user with visual or textual artifacts...The domain of $g$ is $\{0,1\}^{d′}$, i.e. $g$ acts over absence/presence of the interpretable components | This could be a ridge regression model with the form $f(x) \sim x'_1 + x'_2 + \cdots + x'_{d'}$ |
| $\Omega(g)$ | A measure of complexity of the explanation $g\in G$. For example, for decision trees $\Omega(g)$ may be the depth of the tree, while for linear models, $\Omega(g)$ may be the number of non-zero weights. | Seemingly, this is based on how the weights are assigned to the data points. |
| $\pi_x(z)$ | A proximity measure between an instance $z$ to $x$, so as to define locality around $x$. | This could be the Gower distance. |
| $\mathcal{L}(f,g,\pi_x)$ | A measure of how unfaithful $g$ is in approximating $f$ in the locality defined by $\pi_x$. | <span style="color:blue"> Does the LIME R package ever consider this? </span> |

**Goal of LIME**: In order to ensure both interpretability and local fidelity, we must minimize $\mathcal{L}(f,g,\pi_x)$ while having $\Omega(g)$ below enough to be interpretable by humans. The explanation produced by LIME is obtained by the following:
  $$\xi(x) = \underset{g∈G}{\arg\min} \ \mathcal{L}(f,g,\pi_x) + \Omega(g)$$
This  formulation  can  be  used  with  different  explanation families $G$, fidelity functions $\mathcal{L}$, and complexity measures $\Omega$. Here we focus on sparse linear models as explanations, and on performing the search using perturbations.

**Procedure**: We want to minimize the locality-aware loss $\mathcal{L}(f,g, \pi_x)$ without making any assumptions about $f$, since we want the explainer to be model-agnostic.

1. In order to learn the local behavior of $f$ as the interpretable inputs vary, we approximate $\mathcal{L}(f,g,\pi_x)$ by drawing samples, weighted by $\pi_x$. 
2. We sample instances around $x'$ by drawing nonzero elements of $x'$ uniformly at random (where the number of such draws is also uniformly sampled).
3. Given a perturbed sample $z'\in\{0,1\}^{d′}$ (which contains a fraction of the nonzero elements of $x'$), we recover the sample in the original representation $z\in\mathbb{R}^d$ and obtain $f(z)$, which is used as a label for the explanation model. 
4. Given this data set $\mathcal{Z}$ of perturbed samples with the associated labels, we optimize $\mathcal{L}(f,g,\pi_x) + \Omega(g)$ to get an explanation $\xi(x)$.

It is worth noting that our method is fairly robust to sampling noise since the samples are weighted by $\pi_x$.

### R Package (Draft)

@@

# Inputs for LIME Functions

@@

### `lime` function

@@

### `explain` function

>>>>

---
title: "Lime Algorithm and Proposed Methods"
author: "Katherine Goode"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

@@

This journal contains a description of the LIME procedure from the original paper, my write-up of the LIME algorithm used in the R package, and ideas for different binning methods that may be better than the ones currently implemented in the R package.

```{r}
# Source functions
source("../../code/helper_functions.R")
```

# Original Paper

Here, I have copied in the section of the original paper that describes LIME and put it in a more orgainzed format. I have also added in some comments explaining my interpretation of the text and questions where I am not sure what is meant by the text.

### Notation

@@

| Notation | Original Paper Definition | My Interpretation (in terms of a data table) |
| -------- | ------------------------- | ------------------------- |
| $x\in\mathbb{R}^d$ | Original representation of an instance being explained | Vector of observed features for one observation e.g. $(x_1, x_2,...,x_d)$ where $\textbf{X}$ is an $n$ by $d$ matrix | 
| $f:\mathbb{R}^d→\mathbb{R}$ | The model being explained...In classification, $f(x)$ is the probability (or a binary indicator) that $x$ belongs to a certain class. | This is a random forest in our case. |
| $x'\in\{0,1\}^{d'}$ | A binary vector for interpretable representation (where interpretable explanations are defined as "use a representation that is understandable by humans") | Varies based on the simulation method used - note that $1\le d' \le d$ since it is possible that feature selection may be performed to choose a smaller set of variables than originally in $x$ (One example: vector of indicator variables that indicate whether the observed value falls in the same bin as the instance being explained e.g. $x'_i = I[x_i\in \mbox{same variable } i \mbox{ bin as the instance being explained}]$) | In our example, this is the random forest model. |
|$G$ | A class of potentially interpretable models, such as linear models, decision trees, or falling rule lists |  <span style="color:blue"> I'm not sure what is considered $G$ in the R package - the fitting of the final ridge regression model or the ridge regression model used for feature selection? </span>|
| $g$ | An explanation as a model...can be readily presented to the user with visual or textual artifacts...The domain of $g$ is $\{0,1\}^{d′}$, i.e. $g$ acts over absence/presence of the interpretable components | This could be a ridge regression model with the form $f(x) \sim x'_1 + x'_2 + \cdots + x'_{d'}$ |
| $\Omega(g)$ | A measure of complexity of the explanation $g\in G$. For example, for decision trees $\Omega(g)$ may be the depth of the tree, while for linear models, $\Omega(g)$ may be the number of non-zero weights. | Seemingly, this is based on how the weights are assigned to the data points. |
| $\pi_x(z)$ | A proximity measure between an instance $z$ to $x$, so as to define locality around $x$. | This could be the Gower distance. |
| $\mathcal{L}(f,g,\pi_x)$ | A measure of how unfaithful $g$ is in approximating $f$ in the locality defined by $\pi_x$. | <span style="color:blue"> Does the LIME R package ever consider this? </span> |
| $z'\in\{0,1\}^{d′}$ | A perturbed sample...(which contains a fraction of the nonzero elements of $x′$) | <span style="color:blue"> It appears that they are creating the perturbations in a different way than the R package. </span> |
| $z\in\mathbb{R}^d$ | The sample in the original representation | <span style="color:blue"> Again, this seems to be different than the R package. </span> |
| $f(z)$ | Used as a label for the explanation model | This is the application of the complex model to the perturbed sample. |
| $\mathcal{Z}$ | Data set...of perturbed samples with the associated labels | This is the full simulated dataset of perturbations and complex model predictions. |
| $\xi(x)$ | The explanation produced by LIME...$\underset{g∈G}{\arg\min} \ \mathcal{L}(f,g,\pi_x) + \Omega(g)$ | The output from the explanation model which is found through an optimiation process. |

### Goal of LIME

In order to ensure both interpretability and local fidelity, we must minimize $\mathcal{L}(f,g,\pi_x)$ while having $\Omega(g)$ below enough to be interpretable by humans. The explanation produced by LIME is obtained by the following:
  $$\xi(x) = \underset{g∈G}{\arg\min} \ \mathcal{L}(f,g,\pi_x) + \Omega(g)$$
This  formulation  can  be  used  with  different  explanation families $G$, fidelity functions $\mathcal{L}$, and complexity measures $\Omega$. Here we focus on sparse linear models as explanations, and on performing the search using perturbations.

### Procedure

We want to minimize the locality-aware loss $\mathcal{L}(f,g, \pi_x)$ without making any assumptions about $f$, since we want the explainer to be model-agnostic.

| Step Number | Paper Description | My Interpretation |
| :----: | ----------------- | ----------------- |
| 1 | In order to learn the local behavior of $f$ as the interpretable inputs vary, we approximate $\mathcal{L}(f,g,\pi_x)$ by drawing samples, weighted by $\pi_x$. | <span style="color:blue"> I don't understand how drawing these samples approximates $\mathcal{L}(f,g,\pi_x)$ </span> |
| 2 | We sample instances around $x'$ by drawing nonzero elements of $x'$ uniformly at random (where the number of such draws is also uniformly sampled).| <span style="color:blue"> What is meant by drawing elements of $x'$? Is this different than what the R package is doing? </span> I'm wondering if the idea is that $x'=(x'_1, x'_2,...,x'_{d'})=(1, 1, ...., 1)$, and then a sample is taken such as $z'=(0, x'_2, 0, 0, x'_5,...,x'_{d'}) =(0, 1, 0, 0, 1, ..., 1)$. <span style="color:blue"> However, this doesn't really make sense in terms of why $d'$ is different from $d$ </span> | 
| 3 | Given a perturbed sample $z'\in\{0,1\}^{d′}$ (which contains a fraction of the nonzero elements of $x'$), we recover the sample in the original representation $z\in\mathbb{R}^d$ and obtain $f(z)$, which is used as a label for the explanation model. | <span style="color:blue"> How is the original sample recovered? </span> Maybe based on my example in the last step $z=(0, x_2, 0, 0, x_5, ..., x_d)$|
| 4 | Given this data set $\mathcal{Z}$ of perturbed samples with the associated labels, we optimize $\mathcal{L}(f,g,\pi_x) + \Omega(g)$ to get an explanation $\xi(x)$. | <span style="color:blue"> Where is this done in the R package? </span> |

It is worth noting that our method is fairly robust to sampling noise since the samples are weighted by $\pi_x$. <span style="color:blue"> Okay - but this will really depend on how $\pi_x$ is chosen </span>

...let $G$ be the class of linear models, such that $g(z′)=w_g\cdot z′$. <span style="color:blue"> (What is $w_g$? Is this suppose to be a linear model where $w_g$ are the coefficients?) </span> We use the locally weighted square loss as $\mathcal{L}$...where we let 
  $$\pi_x(z)=exp(−D(x,z)^2/\sigma^2)$$ 
be an exponential kernel defined on some distance function $D$ (e.g. cosine distance for text, $L2$ distancef or images) with width $\sigma$.
  $$\mathcal{L}(f,g,\pi_x)=\sum_{z, z'\in\mathcal{Z}}\pi_x(z)\left(f(z)-g(z')\right)^2$$

### Comment on Faithfulness

...our choice of $G$ (sparse linear models) means that if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation.  However, we can estimate the faithfulness of the explanation on $\mathcal{Z}$, and present this information to the user. This estimate of faithfulness can also be used for selecting an appropriate family of explanations from a set of multiple interpretable model classes, thus adapting to the given dataset and the classifier. We leave such exploration for future work, as linear explanations work quite well for multiple black-box models in our experiments.

# R Package

### Procedure (Draft)

@@

### Function Inputs

@@

##### `lime` function

@@

##### `explain` function

### 2019-08-13 papers/lime_diagnostics/code/figure-creation.Rmd
work from Tuesday before meeting

>>>>

# Second Attempt

```{r}
# Generate data
set.seed(20190624)
all_observations <- data.frame(case = factor("Other", levels = c("Other", "Prediction of Interest")),
                        feature1 = sort(runif(250, 0, 1)),
                        feature2 = sample(x = 1:250, size = 250, replace = FALSE)) %>%
  mutate(prediction = if_else(feature1 >= 0 & feature1 < 0.1, (0.3 * feature1) + rnorm(n(), 0, 0.01), 
                      if_else(feature1 >= 0.1 & feature1 < 0.3, rbeta(n(), 1, 0.5), 
                      if_else(feature1 >= 0.3 & feature1 < 0.5, sin(pi* feature1) + rnorm(n(), 0, 0.5),
                      if_else(feature1 >= 0.5 & feature1 < 0.8, -(sin(pi* feature1) + rnorm(n(), 0, 0.1)) + 1,
                      if_else(feature1 >= 0.8 & feature1 < 0.9, 0.5 + runif(n(), -0.5, 0.5), 
                              0.5 + rnorm(n(), 0, 0.3)))))))
```

```{r}
# Specify a prediction of interest
prediction_of_interest <- data.frame(case = factor("Prediction of Interest", 
                                                   levels = c("Other", "Prediction of Interest")),
                                     feature1 = 0.07,
                                     feature2 = 200,
                                     prediction = 0.05)
```

```{r}
lime_data <- bind_rows(all_observations, prediction_of_interest)
```

```{r}
# Compute the distance between the prediction of interest 
lime_data$distance <- (1 - gower_dist(x = prediction_of_interest, y = lime_data))^20
lime_data$distance_bad <- (1 - gower_dist(x = prediction_of_interest, y = lime_data))^1

# Fit the interpretable explainer models
lime_explainer <- lm(prediction ~ feature1 + feature2, data = lime_data, weights = distance)
lime_explainer_bad <- lm(prediction ~ feature1 + feature2, data = lime_data, weights = distance_bad)
```

```{r}
lime_data_gathered <- lime_data %>%
  gather(key = feature, value = value, 2:3) %>%
  gather(key = distance_case, value = distance_value, 3:4)
```

```{r}
ggplot(data = lime_data_gathered, 
       mapping = aes(x = value, 
                     y = prediction, 
                     size = distance_value,
                     shape = case)) + 
  geom_point(alpha = 0.5) + 
  geom_abline(data = lime_data_gathered %>% 
                filter(feature == "feature1", distance_case == "distance"),
              intercept = coef(lime_explainer)[[1]] +
                coef(lime_explainer)[[3]]*prediction_of_interest$feature2, 
              slope = coef(lime_explainer)[[2]], 
              size = 1) +
  #stat_smooth(data = lime_data_gathered %>% filter(feature == "feature2")) +
  facet_grid(distance_case ~ feature, scales = "free") + 
  theme_bw()
```

### 2019-08-13 papers/lime_diagnostics/visual-diagnostics-of-a-model-explainer.Rnw
work from Tuesday before meeting

\begin{document}

@@

In the field of statistics, there are two main uses for models: inference and prediction. Machine learning models are often used for the latter purpose. While these models have been proven to perform well in a wide range of prediction problems, their accuracy comes at the cost of interpretability. They are often complex algorithms that are good at identifying patterns in the data but \hh{often} lack a functional form, \hh{making it impossible} to directly interpret them. As a result, they \hh{HH: who is they - make that explicit. always.} produce predictions that lack an explanation, which has earned them the reputation of being ``black-box models". 
\hh{HH: I would make the claim that the need for explanations goes even deeper. Even if a functional form exists, such as in a linear model, there are situations where the model gets complicated enough to be not or not easily interpretable. Only very good natured three-way interactions in a logistic regression are interpretable. For the other ones more work is needed than most audiences would be capable of. Explainer models can be used in those situations as well.}

@@

@@

The current implementations of LIME use a linear regression model as the explainer model \citep{lime} (references to R and Python packages) \hh{for which models? LIME uses neural networks for image analyses}. These implementations are relying on the assumption that the relationship between the complex model predictions and the features is linear in a local region. \hh{HH: be a bit careful in the phrasing here - we can always approximate any function using a linear form (think Taylor expansion). The question is how big the error is. } It is important to assess this assumption in the local region of interest in order to trust the explanations produced by LIME. \hh{The next sentence is muddying the waters - by being able to doagnose explainer models we will be able to assess the different settings offered in the implementations. Let's bring those up later. }
Additionally, the implementations offer various input settings, but little work has been done to provide advice on how to specify the settings in practice (look more into this). An assessment of the explanations could also help to determine which settings produce the explainer model with the best approximation of the complex model.

>>>>

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

In the field of statistics, there are two main uses for models: inference and prediction. Machine learning models are often used for the latter purpose. While these models have been proven to perform well in a wide range of prediction problems, their accuracy comes at the cost of interpretability. Machine learning models are generally complex algorithms that are good at identifying patterns in the data but \hh{often} lack a functional form, which makes it impossible to directly interpret them \kg{***Should I change this them?}. As a result, machine learning learning models produce predictions that lack an explanation, which has earned them \kg{***What about this one?} the reputation of being ``black-box models". 
\hh{HH: I would make the claim that the need for explanations goes even deeper. Even if a functional form exists, such as in a linear model, there are situations where the model gets complicated enough to be not or not easily interpretable. Only very good natured three-way interactions in a logistic regression are interpretable. For the other ones more work is needed than most audiences would be capable of. Explainer models can be used in those situations as well.}

@@

\kg{While reading through the original paper again, I rediscovered this quote. This may be something to reference or quote in this section: ``...our choice of $G$ (sparse linear models) means that if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation. However, we can estimate the faithfulness of the explanation on $\mathcal{Z}$, and present this information to the user. This estimate of faithfulness can also be used for selecting an appropriate family of explanations from a set of multiple interpretable model classes, thus adapting to the given dataset and the classifier. We leave such exploration for future work, as linear explanations work quite well for multiple black-box models in our experiments."}

@@

The current implementations of LIME use a linear regression model as the explainer model \citep{lime} (references to R and Python packages) \hh{for which models? LIME uses neural networks for image analyses} \kg{I think that it is actualy using ridge regression for all cases - even images}. These implementations are relying on the assumption that the relationship between the complex model predictions and the features is linear in a local region. \hh{HH: be a bit careful in the phrasing here - we can always approximate any function using a linear form (think Taylor expansion). The question is how big the error is. } It is important to assess this assumption in the local region of interest in order to trust the explanations produced by LIME. \hh{The next sentence is muddying the waters - by being able to doagnose explainer models we will be able to assess the different settings offered in the implementations. Let's bring those up later. }
Additionally, the implementations offer various input settings, but little work has been done to provide advice on how to specify the settings in practice (look more into this). An assessment of the explanations could also help to determine which settings produce the explainer model with the best approximation of the complex model.
